Vertical shear instability in astrophysical disks: vertically global linear calculations with thermal relaxation and applications to the Solar Nebula

### 2015-03-30 summary.tex
waiting for zmax3 max rate fig

\section{Summary and discussion}\label{summary}
In this paper we have studied the  axisymmetric vertical shear
instability (VSI) in astrophysical disks that arise from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. We examined how
the linear instability depends on disk thermodynamics and applied our
results to assess the relevance of the VSI in realistic protoplanetary
disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14}, which assumed isothermal
perturbations, to include an energy equation. Our effort may also be
regarded as an extension of \cite{urpin03}, which did include an
energy equation, but adopted the Boussinesq and vertically local
approximations. We considered vertically global and fully
compressible perturbations.

We included a source term in the energy equation which relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach allows the thermodynamic
response of the disk to be controlled by varying the parameter
$\beta$, and has been employed in previous numerical simulations of
the VSI \citep{nelson13}. However, it is possible to 
relate  $\beta$ to physical thermal timescales expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. For
analytic discussion only, we further simplified the linear problem by
ignoring the background radial disk structure. This is equivalent to adopting
the `vertically global shearing box' (VGSB) formalism developed by
\cite{mcnally14}. 
We find the VGSB framework is appropriate for analyzing the VSI because 
rapid thermal relaxation ($\beta\ll1$) is 
generally required for the instability.   However, the VGSB framework is not
appropriate in the adiabatic limit ($\beta \gg 1$).

We mostly focused on vertically isothermal disks ($\Gamma=1$) as 
this is the relevant case for protoplanetary disks that are passively irradiated by its central star 
\citep{chiang97}. It also permits a largely analytical discussion.
Our analysis is an extension of the linear models considered in
\cite{lubow93} by  adding vertical shear and (rapid) thermal
relaxation. We obtained explicit solutions describing the VSI with 
thermal relaxation in the thin disk limit.

The fundamental VSI mode is characterized by a symmetric vertical
velocity perturbation with zero nodes and $\delta
v_z\sim\mathrm{constant}$ near the mid-plane, and by an anti-symmetric
density perturbation with one node at $z=0$.  These have been referred
to as `corrugation modes' by \cite{nelson13}, and were found to
dominate their numerical simulations, as well as that of
\cite{stoll14}. Furthermore, when computing unstable modes for a
protoplanetary disk model with thermal relaxation based on realistic
disk properties, we find the fundamental mode to be the dominant
mode. The fundamental mode is
thus likely the most relevant VSI mode in protoplanetary disks.

We determined the critical thermal relaxation timescale 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$, below which small-scale
perturbations may be regarded as isothermal and the VSI 
operates. We have therefore provided a quantitative measure for the `fast'
thermal relaxation requirement for the VSI in astrophysical disks.

We also solved the linear stability problem numerically using a
pseudo-spectral approach. For numerical calculations we relax all the
approximations made in our analysis, and accounted for
the background radial disk structure.  Our
numerical results are mostly consistent with our analytic 
expectations, including the dependence of $\beta_\mathrm{crit}$ on
disk parameters.  We find the fundamental mode is robust against
boundary conditions, and higher order modes are more rapidly
stabilized with increasing $\beta$.

We briefly studied vertically non-isothermal disks. We confirm the finding by \cite{nelson13} that in the absence
of vertical entropy gradients, the VSI does not require rapid thermal
relaxation.  When a stable stratification is introduced, the
rapid reduction in VSI growth rates by a finite thermal relaxation
timescale is qualitatively similar to that observed for vertically
isothermal disks. Based on numerical results, we conjecture that the
critical thermal relaxation timescale is given by 
$\beta_\mathrm{crit}=\epsilon|q|/(\gamma-\Gamma)$ in vertically 
non-isothermal disks.

%application to MMSN

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal relaxation timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a useful first step to estimate the relevance of the VSI in
realistic astrophysical disks without resorting to non-linear
numerical simulations.

@@

We have neglected viscosity. \cite{nelson13} find that an 
alpha viscosity $\lesssim O(10^{-4})$ is required for the VSI. 
It is of interest to quantify the effect of viscosity 
on the linear VSI. This would help identify, more accurately, the
appropriate regions in a protoplanetary disk that are sufficiently
laminar to develop the VSI.  This problem is, however, complicated by
the fact that proper inclusion of viscosity introduces a meridional
flow in the equilibrium state, as well as viscous heating
in the energy equation.

>>>>

\section{Summary and discussion}\label{summary}
In this paper we have studied the axisymmetric vertical shear
instability (VSI) in astrophysical disks that arise from the
height dependence of the equilibrium angular velocity profile
$\Omega=\Omega(r,z)$ in a baroclinic disk. Vertical shear is generally
expected in astrophysical disks due to the radial dependence of the
disk temperature, or more generally the entropy. However, the VSI
requires a short thermal timescale in sub-adiabatically stratified
disks, which is also typical of astrophysical disks. Physically, a
short thermal timescale is needed to overcome the stabilizing
influence of buoynancy forces.

We thus examined how the linear VSI depend on disk
thermodynamics. Our primary goal was to quantify a characteristic
thermal timescale below which the VSI can operate in stably stratified
disks, thereby provide a simple way to assess whether or not the VSI
is relevant in realistic disks.

We generalized previous vertically global linear 
calculations of the VSI \citep{nelson13,mcnally14,barker15}, which
assumed isothermal perturbations, to include an energy equation with % Our 
% effort may also be regarded as an extension of the original VSI study
% by \cite{urpin03} who did account for thermal evolution, but adopted 
% the Boussinesq and vertically local approximations. We considered
% vertically global and fully compressible perturbations. 
a source term in that relaxes the disk
temperature to its initial value on a timescale
$t_c=\beta\Omega_k^{-1}$. This approach allows the thermodynamic
response of the disk to be controlled by varying the parameter
$\beta$, and has been employed in previous numerical simulations of
the VSI \citep{nelson13}. However, it is possible to 
relate  $\beta$ to physical thermal timescales expected in realistic disks.

In accordance with the above studies, which show that the VSI occurs for 
perturbations with small radial lengthscales, we applied the 
radially local approximation in the linearized fluid equations. % For
% analytic discussion only, we further simplified the linear problem by
% ignoring the background radial disk structure. This is equivalent to adopting
% the `vertically global shearing box' (VGSB) formalism developed by
% \cite{mcnally14}. 
% We find the VGSB framework is appropriate for analyzing the VSI because 
% rapid thermal relaxation ($\beta\ll1$) is generally required for the
% instability to operate in stably stratified disks 
% However, the VGSB framework is not
% appropriate in the adiabatic limit ($\beta \gg 1$). 
We mostly focused on vertically isothermal disks ($\Gamma=1$) as 
this is the relevant case for protoplanetary disks that are passively
irradiated by its central star \citep{chiang97}. It also permits a
largely analytical discussion. Our analysis is an extension of the
linear models considered in \cite{lubow93} by adding vertical shear
and (rapid) thermal relaxation. We obtained explicit solutions
describing the VSI with thermal relaxation in the thin disk limit.

% The fundamental VSI mode is characterized by a symmetric vertical
% velocity perturbation with zero nodes and $\delta
% v_z\sim\mathrm{constant}$ near the mid-plane, and by an anti-symmetric
% density perturbation with one node at $z=0$.  These have been referred
% to as `corrugation modes' by \cite{nelson13}, and were found to
% dominate their numerical simulations, as well as that of
% \cite{stoll14}. Furthermore, when computing unstable modes for a
% protoplanetary disk model with thermal relaxation based on realistic
% disk properties, we find the fundamental mode to be the dominant
% mode. The fundamental mode is
% thus likely the most relevant VSI mode in protoplanetary disks.

For a vertically isothermal disk we have determined a thermal
relaxation timescale $\beta_\mathrm{crit}=\epsilon|q|/(\gamma-1)$,
below which perturbations may be regarded as isothermal and the VSI 
operates effectively. While $\beta_\mathrm{crit}$ was derived for the
lowest order VSI mode --- the `fundamental 
corrugation mode' --- we find from numerical solutions to the  linear
problem that $\beta_\mathrm{crit}$ in fact provides a useful way to
characterize the stabilization of the VSI by a finite thermal 
timescale. That is, all VSI growth rates are reduced significantly as
$\beta$ is increased from zero to $\beta_\mathrm{crit}$.  
% This is despite the fact the occurence of `surface
% modes' associated with imposed vertical boundaries in numerical
% models of vertically isothermal disks. These modes are absent in our
% analysis since the vertically isothermal disk has no surface. 
We have therefore provided a quantitative measure for the 
`short' thermal timescale requirement for the VSI in astrophysical
disks.

%Despite complications from imposing a vertical boundary in numerical
%calculations in a vertically isothermal disk, which gives rise to
%surface modes \citep{barker15} (and other boundary effects).

% We also solved the linear stability problem numerically using a
% pseudo-spectral approach. For numerical calculations we relax all the
% approximations made in our analysis, and accounted for
% the background radial disk structure.  Our
% numerical results are mostly consistent with our analytic 
% expectations, including the dependence of $\beta_\mathrm{crit}$ on
% disk parameters.  
% We find the fundamental mode is robust against
% boundary conditions, and higher order modes are more rapidly
% stabilized with increasing $\beta$. 
% We briefly studied vertically non-isothermal disks. We confirm the finding by \cite{nelson13} that in the absence
% of vertical entropy gradients, the VSI does not require rapid thermal
% relaxation.  When a stable stratification is introduced, the
% rapid reduction in VSI growth rates by a finite thermal relaxation
% timescale is qualitatively similar to that observed for vertically
% isothermal disks. Based on numerical results, we conjecture that the
% critical thermal relaxation timescale is given by 
% $\beta_\mathrm{crit}=\epsilon|s|/(\gamma-\Gamma)$ in vertically 
% non-isothermal disks.  
%application to MMSN

We applied our linear models to assess the applicability of the VSI in
protoplanetary disks. We considered the Minimum Mass Solar Nebula 
(MMSN) described in \cite{chiang10} and   
modeled the thermal timescale $\beta$ based on realistic
disk properties and dust opacity. This results in a $\beta$ function that
depends on both the perturbation lengthscale as well as the height 
from the midplane. We find the MMSN is unstable to the VSI 
from $\sim 5$AU to $\sim50$AU with characteristic growth times of
$\sim 30$ orbits. While it is possible to have the VSI at even
smaller radii, this would require very small radial lengthscales, 
which may be subject to viscous decay in a real disk. 
%or  
% higher growth rates by having
% surface modes --- these are dependent
% on boundary conditions in the
% vertically isothermal disk

Although $\beta$ is not vertically constant in the
MMSN as assumed in our analysis, we find that evaluating the condition
$\beta <\beta_\mathrm{crit}$ at the disk midplane gives a rough measure of the
radii at which the VSI operates in the MMSN. This simple criterion is
thus a practical first step in estimating the relevance of the VSI in
realistic astrophysical disks without resorting to detailed linear
analysis or non-linear numerical simulations.

@@

We have neglected viscosity. \cite{nelson13} find that an alpha
viscosity $\lesssim O(10^{-4})$ is required for the
VSI. \cite{barker15} found that inclusion of viscosity is needed for 
convergence when studying surface modes. Viscosity may also affect the
thermal timescale requirement of the VSI, because viscous heating
needs to be self-consistently included in the energy equation. This
should be investigated in future works.

% It is of interest to quantify the effect of viscosity 
% on the linear VSI. This would help identify, more accurately, the
% appropriate regions in a protoplanetary disk that are sufficiently
% laminar to develop the VSI.  This problem is, however, complicated by
% the fact that proper inclusion of viscosity introduces a meridional
% flow in the equilibrium state, as well as viscous heating
% in the energy equation.
